---
title: "strawberries report"
authors:  
format: html
editor: visual
---

# Strawberries

```{r}
#| label: load libraries and set options 
#| warning: false
#| message: false
#| echo: false

library(knitr)  
library(kableExtra)
library(tidyverse)
library(stringr)
library(labelled)
library(data.table)
library(scales)

source("Functions.R")
```

## Introduction

Strawberries are one of the most widely cultivated fruit crops in the United States, with production heavily concentrated in states such as California and Florida. To sustain high yields and consistent quality, strawberry growers depend on fertilizers to supply essential plant nutrients. However, fertilizer use also raises important questions about environmental impact, soil health, and sustainability.

When you go the market and look in the produce section, you see all sorts of fruits and vegetables, but do you ever stop to think about where they are coming from or the farming process that got them there for you to purchase and consume? Here we will specifically look at the production of strawberries in the United States.

This project analyzes data on strawberry fertilizer use across US states. The United States Department of Agriculture (USDA) collects census and survey data regularly on all of the major crops grown in the United States. To analyze the strawberry crop, we will look at data collected by the USDA through census' and surveys from 1997 to 2025. We will focus on the six states that produce the majority of the strawberries consumed: California, Florida, Washington, Oregon, North Carolina and New York. The goal is to explore patterns, trends, and potential pros/cons of different fertilizer type including Nitrogen, Phosphate, Potash and Sulfur to understand how these inputs vary by geography and time.

Main data source: [USDA Quickstats](https://quickstats.nass.usda.gov/)

We begin by cleaning and structuring the data to ensure technical correctness, then conduct exploratory analyses that focus on:

-   Usage distribution by fertilizer type

-   Geographic difference in fertilizer use

-   Trends in fertilizer use over time

-   Relationships among fertilizer types

The approach follows a structured analytically strategy

1.  Clean → 2. Explore → 3. Interpret → 4. Report

The purpose of the exploratory data analysis is to not draw conclusion but to hopefully have a better understanding on how fertilizer use reflects broader management practices in strawberry agriculture. The results can serve as a foundation for future work linking fertilizer management to yield, soil health and environmental sustainability.

## Raw Data and Acquisition

The data set was provided by the USDA Strawberry Production Survey. It contains information about fertilizer application types, quantities, and related descriptors by US states and year.

Fertilizer use is a key input determining yield and environmental impact. Comparing fertilizer types (organic vs conventional) provides insight into production sustainability. We are looking at a comparison between organic and conventional strawberries. Fertilizers used in the strawberries are used to determine organic and conventional products

| Chemical | Role | Typical Form | Organic Status |
|------------------|------------------|------------------|------------------|
| Nitrogen (N) | Leaf/vegetative growth | Urea, Ammonium Nitrate | Not organic unless it came from compost |
| Phosphate (P) | Root development | Superphosphate | Synthetic |
| Potash (K) | Disease resistance | Potassium chlroride or Sulfate | Natural forms like potassium from mineral sources can be organic |
| Sulfur (S) | Disease control | Elemental sulfur | Allows in orgnanic systems if naturally derived |

To achieve a technically correct dataset, we implemented a modular cleaning pipeline that

-   Standardized column names

-   Trimmed white spaces and punctuation

-   Removed uninformative columns (state_ansi, constant columns)

-   Passed numeric columns safely

-   Cleaned text fields (data_item, domain_category -\> fertilizer)

-   Removed NA and zero values

## What fertilizers are being used in strawberry farming?

Fertilizer is an important element in the farming process because, just like humans, plants need essential nutrients to thrive. Think of fertilizer like a multivitamin for humans. The most important nutrients for plants are carbon, hydrogen, and oxygen. Plants get these from the air and water. The others are obtained from soil, but sometimes the soil isn't naturally nutrient-dense or has been farmed many times, and the plants need supplemental nutrients. The primary macro nutrients plants need are nitrogen, phosphorus, and potassium. The secondary macro nutrients for plants are sulfur, calcium, and magnesium. Plants also require further micro nutrients, but for this research, we will focus on a few of the macro nutrients.

We sourced some data to analyze the use of fertilizer on strawberry crops from the same USDA database above. The four nutrient fertilizers that the USDA collected data about are: Nitrogen, Phosphate, Potash, and Sulfur. Nitrogen is helpful to support green, leafy growth. Phosphate helps the root, flower, and fruit development. Potash (or potassium) helps a plant's stress tolerance and disease resistance. Sulfur helps overall plant health and enzyme and nutrient metabolism. These are important macro nutrients for healthy plant growth, along with others.

## What constitutes organic produce?

For a crop to be considered organic, there are certain conditions that need to be followed by the farmers. This means farmers cannot use certain typical farming practices used to grow conventional products. Farmers cannot use inorganic fertilizers, genetic engineering, or radiation on their products.

For a product to have the USDA Organic seal, the product must have more than 95% certified organic ingredients. One of the biggest differences is that organic produce is exposed to fewer toxic pesticides and chemicals during the growing process.

## Are these fertilizers used on conventional and/or organic strawberries?

Since all four of these nutrients are essential for crop growth, they are used by farmers growing conventional and organic strawberries. However, the source of the fertilizer differs because there are organic and inorganic fertilizers.

Inorganic fertilizers are typically manufactured, water-soluble products and are not carbon-based. They generally tend to be cheaper due to their manufacturing process and have easy calculations to determine the application rate. Depending on the form of the fertilizer, you can also buy liquid fertilizer for quick absorption or slow-release forms of fertilizer for a release of nutrients over time. One of the negatives to inorganic fertilizers is that some of the synthetic compounds used can cause environmental harm through leaching or runoff.

Organic fertilizers are typically derived from living organisms and are carbon-based. Some fertilizers might be derived from fish emulsion or blood meal as the base. Since they are derived from these living organisms, organic fertilizers are generally more expensive and considered to be slow-release. This is because the soil needs a couple of days to weeks to transform the organic nutrients into a form the plants can more readily absorb. Because of this, it can be difficult to calculate application rates with precision. One of the negatives of organic fertilizer is that it may contain weeds or pathogens, like Salmonella, since both can be found in the various living organisms that the fertilizer is derived from.

While the choice of fertilizer is important, it is still recommended to add other organic materials to help support the soil, specifically carbon-based materials like compost or grass clippings. Items like these or other ground cover can help plants receive the other macro and micro nutrients needed for healthy growth.

## Topic: Explore Fertilizer Use in Strawberry Cultivation.

Exploratory Questions:

1.  Why is fertilizer important?

Nitrogen fertilizer was created by Sam Eyde and Kristian Birkeland in December of 1905. This has changed agriculture and contributed to the high crop yields of today. The possibility of feeding the growing population of the world is thanks to fertilizer improvements. The critical plant nutrients are the macro nutrients nitrogen, phosphorus, potassium, calcium, sulphur, magnesium, oxygen, hydrogen, carbon, and the micro nutrients iron, boron, chlorine, manganese, zinc, copper, molybdenum, and nickel. Planting crops depletes the soil, and the fertilizers replenish it to ensure that future crops have a good foundation for growth.

2.  What are the most used fertilizers by state, based on the data set?

From the fertilizer_state data, we can see that many fertilizers are used equally as often. Nitrogen, Phosphate, and Potash (Potassium) seem to be used an equal number of times across the different states in the data set. This aligns with the information from the Yara website.

## Data Wrangling Process

#### Cleaning the Conventional Strawberry Data Set

In order to work with the data, we must read the raw csv file from the USDA into R. Here is a glimpse of the raw data set: <br>

```{r}
#| label: read in the data & run drop 1-item columns function
#| echo: false

# Read in the csv data file
strawb <- read_csv("strawberry_10Oct25.csv", col_names = TRUE)

# Run the function on the strawberry data
strawb <- drop_one_value_col(strawb)
```

<br> Now that we've taken a glance at the data, the first step is to remove any columns that are entirely one data value and thus will not be useful to our analysis. This removal of unnecessary columns also makes the rest of the data easier to manage. Next, we analyze what we have left. Upon looking at the data, we have a couple columns of numeric data that are currently character structure. In order to be able to work with the data and perform exploratory analysis, we must make sure the structure of all numeric columns are in the correct format for analysis and manipulation, so we made those changes.

```{r}
#| label: Verify the class of columns 
#| warning: false
#| echo: false

# Converting the Value and CV(%) columns from character into numeric structure.
strawb <- strawb %>% 
  mutate(Value = as.numeric(gsub(",", "", strawb$Value))) %>% 
  mutate(`CV (%)` = as.numeric(strawb$`CV (%)`))

```

Next, we investigated the `Data Item` column. This column gives us a lot of different information all in one cell. In order to make it more useful, we will remove the phrases common to all elements of the data set. For example, removing the "strawberries - " from the beginning of each line. Thus, it becomes a more informative column about the numeric data values.

```{r}
#| label: Clean up Data Item column
#| echo: false

# Remove strawberries from the start of every line in Data Item
strawb$`Data Item` <- str_replace(strawb$`Data Item`, "^ST.*IES((\ -\ )|(,\ ))", "")

```

In this data set, the data comes from surveys and census'. There are some differences in the data collection for each of those methods that will make it difficult to clean the data set while it is all together. So next, we will split the data into separate survey and census data frames for further cleaning. We also have a separate data set for the organic data.

```{r}
#| label: split survey/census & drop all one again
#| echo: false

# Filter for only the SURVEY data
strawb_survey <- strawb %>% 
  filter(Program == "SURVEY")

# Filter for only the CENSUS data
strawb_census <- strawb %>% 
  filter(Program == "CENSUS")

# Run drop one value columns again
strawb_survey <- drop_one_value_col(strawb_survey)
strawb_census <- drop_one_value_col(strawb_census)

```

#### Cleaning the Census Portion

Once we've split the data, the census data didn't need any further cleaning. We also pivoted it to it's wider form to be able to see the data is a more tabular format. In the wide format, each row represents the year's output for a state and each data observation for that year is in a column.

```{r}
#| label: pivot strawb_census wider 
#| echo: false

strawb_census_wide <- strawb_census %>%
  pivot_wider(
    id_cols = c(Year, State),          # Columns that uniquely identify each new wide row
    names_from = `Data Item`,          # Column whose values will become the new column names
    values_from = c(Value, `CV (%)`)   # Column whose values will fill the new columns
  )
```

#### Cleaning the Survey Portion

The survey data is a bit more complicated because each state has a bit of variability in how they report the information. We will need to further split up the survey data by state. Recall, the states we are looking at are California, Florida, New York, North Carolina, Oregon, and Washington.

Before making that split, we chose to remove the rows where the data was a forecast value. The rational behind this was that the actual year value was also provided. After that, we split the survey data up into a data frame for each of the six states and look at each separately.

```{r}
#| label: remove "Forecast" rows from the dataset
#| echo: false

# Remove rows predicting "FORCAST"
strawb_survey <- strawb_survey %>% 
  filter(!str_detect(Period, "FORECAST")) 

```

For most of the states, we will remove the `Week Ending` column and the `State ANSI` column, as they would be removed by the drop one value column function. We also wrote and used a function to drop duplicate rows. In some of the states, the survey recorded "marketing year" and "year" with the same value. We did not need to keep both rows as they ended up representing the same value.

Cleaning the data sets for New York, Washington, and Oregon were the easiest. We dropped the unnecessary columns and ran the drop duplicate rows function and finished with a clean data set.

```{r}
#| label: looking at and cleaning New York survey data
#| echo: false

# Filter for NY and remove unneeded columns
new_york <- strawb_survey %>% 
  filter(State == "NEW YORK") %>% 
  select(-`Week Ending`, -`State ANSI`)

# Run function to remove duplicate rows
new_york <- drop_duplicate_rows(new_york)

```

```{r}
#| label: looking at and cleaning Washington survey data
#| echo: false

# Filter for WA and remove unneeded columns
washington <- strawb_survey %>% 
  filter(State == "WASHINGTON") %>% 
  select(-`Week Ending`, -`State ANSI`)

# Run function to remove duplicate rows
washington <- drop_duplicate_rows(washington)

```

```{r}
#| label: looking at and cleaning Oregon survey data
#| echo: false

# Filter for OR and remove unneeded columns
oregon <- strawb_survey %>% 
  filter(State == "OREGON") %>% 
  select(-`Week Ending`, -`State ANSI`) 

# Run function to remove duplicate rows
oregon <- drop_duplicate_rows(oregon)

```

North Carolina was the only state that reported their data in weekly increments rather than yearly. To address that, we took the sum of all of the weeks in the year and now have a value for the year. This will allow us to compare the data with the other states.

```{r}
#| label: looking at and cleaning North Carolina survey data
#| echo: false

# Filter for NC and remove unneeded columns
ncarolina <- strawb_survey %>% 
  filter(State == "NORTH CAROLINA") %>% 
  select(-`State ANSI`)

# Run function to remove duplicate rows
ncarolina <- drop_duplicate_rows(ncarolina)

# Summarize the weekly values to a year average
ncarolina <- ncarolina %>%
  group_by(Year, State, `Data Item`) %>%
  summarize(Value = sum(Value), Period = "YEAR", .groups = "drop") %>% 
  ungroup() %>% 
  # remove the rows with period as previous year because we have them as the actual year
  filter(!str_detect(`Data Item`, "PREVIOUS YEAR,")) %>% 
  relocate(Period, .after = Year)

```

California and Florida began the same way as the rest of the states. They both had extra data on monthly values in addition to yearly values. Since we are mainly looking at yearly values to compare between states, we removed the monthly value rows for better comparisons.

```{r}
#| label: looking at and cleaning California survey data
#| echo: false

# Filter for CA and remove unneeded columns
calif <- strawb_survey %>% 
  filter(State == "CALIFORNIA") %>% 
  select(-`Week Ending`, -`State ANSI`) 

# Run function to remove duplicate rows
calif <- drop_duplicate_rows(calif)

# Filter for year periods
calif <- calif %>% 
  filter(str_detect(Period, "YEAR"))

```

```{r}
#| label: looking at and cleaning Florida survey data
#| echo: false

# Filter for FL and remove unneeded columns
florida <- strawb_survey %>% 
  filter(State == "FLORIDA") %>% 
  select(-`Week Ending`, -`State ANSI`) 

# Run function to remove duplicate rows
florida <- drop_duplicate_rows(florida)

# Filter for year periods
florida <- florida %>% 
  filter(str_detect(Period, "YEAR"))

```

Once each state has a clean data set, we can recombine the data into a single data frame for further exploration and also pivot the data wider for visualization in a more tabular format.

```{r}
#| label: put state survey data back together
#| echo: false

# Create a list containing all of them
states_list <- list(new_york, oregon, ncarolina, calif, florida, washington)

# Bind all the data frames together into a single data table
strawb_survey <- rbindlist(states_list)

# Clean up the environment
rm(states_list)

```

```{r}
#| label: Pivot wider strawb_survey
#| echo: false

strawb_survey_wide <- strawb_survey %>%
  pivot_wider(
    id_cols = c(Year, State),     # Columns that uniquely identify each new wide row
    names_from = `Data Item`,     # Column whose values will become the new column names
    values_from = Value           # Column whose values will fill the new columns
  )


```

#### Cleaning the Organic Strawberry Data Set

First, we read in a separate .csv file for organic strawberry data. Once the data was read in, we took very similar steps to those taken in cleaning the conventional strawberry data set.

```{r}
#| label: read in the organic csv & run drop 1-item columns function on data
#| echo: false

strawb_organic <- read.csv("strawb_organic.csv")
strawb_organic <- drop_one_value_col(strawb_organic)

```

The data had lots of unnecessary columns, so we ran our drop one value columns functions. After that, we made sure to covert the numeric value columns into the correct structure for analysis.

```{r}
#| label: clean up the columns and class of variables.
#| warning: false
#| echo: false

# Remove other unnecessary columns
strawb_organic <- strawb_organic %>% 
  select(-State.ANSI) 

# Converting the Value and CV(%) columns from character into numeric structure.
strawb_organic <- strawb_organic %>% 
  mutate(Value = as.numeric(gsub(",", "", strawb_organic$Value))) %>% 
  mutate(`CV....` = as.numeric(CV....))

```

The last step in getting the organic data clean is to remove the phrase common to the beginning of every row of the `Data Item` column. Once this is removed, the organic data is ready for analysis.

```{r}
#| label: Clean-up data item column
#| echo: false

# Remove strawberries/organic from the start of every line in Data Item
strawb_organic$Data.Item <- str_replace(
  string = strawb_organic$Data.Item,
  pattern = "^(?:STRAWBERRIES, ORGANIC - |STRAWBERRIES, ORGANIC, )",
  replacement = ""
)
```

```{r}
#| label: pivot organic wider 
#| echo: false

strawb_organic_wide <- strawb_organic %>%
  pivot_wider(
    id_cols = c(Year, State),        # Columns that uniquely identify each new wide row
    names_from = Data.Item,          # Columns whose values will become the new column names
    values_from = c(Value, CV....)   # Column whose values will fill the new columns
  )

```

#### Cleaning the Fertilizer Data Set

The fertilizer data again came for the USDA in its own .csv file. We began the same as before, reading in the data, running the drop one value columns function to remove unnecessary columns, and changing the numeric value columns to the correct structure for analysis.

```{r}
#| label: read in the fertilizer data
#| echo: false

# Read in csv file from USDA
strawb_fertilizers <- read.csv("strawberry_fertilizers.csv") 

# Run the function on the strawberry data
strawb_fertilizers <- drop_one_value_col(strawb_fertilizers)
```

```{r}
#| label: Convert Value column to numeric
#| warning: false
#| echo: false

# Convert to numeric
strawb_fertilizers <- strawb_fertilizers %>% 
  mutate(Value = as.numeric(gsub(",", "", Value)))

```

Next, we have a new column compared to the other data sets that tells us the type of fertilizer used. In this column, similar to the `Data Item` column previously, we need to remove some common words to every entry. We will do that to the `Fertilizer` column and again to the `Data Item` column so that it is clean what data value is being collected for each fertilizer type.

Now, all of our data sets are clean and ready for analysis.

```{r}
#| label: Clean up Domain.Category to Fertilizer
#| echo: false

# Rename Fertilizer column and clean it up 
strawb_fertilizers <- strawb_fertilizers %>% 
  rename(Fertilizer = Domain.Category) %>% 
  mutate(Fertilizer = sub(".*\\((.*?)\\).*", "\\1", Fertilizer))

```

```{r}
#| label: Clean up Data.Item column
#| echo: false

# Remove strawberries from the start of every line in Data Item
strawb_fertilizers <- strawb_fertilizers %>% 
  mutate(Data.Item = str_remove(Data.Item, ".*-")) 

```

```{r}
#| label: pivot fertilizer wider 
#| echo: false

strawb_fertilizers_wide <- strawb_fertilizers %>%
  pivot_wider(
    id_cols = c(Year, State),  # Columns that uniquely identify each new wide row
    names_from = c(Fertilizer, Data.Item), # Columns whose values will become the new column names
    values_from = Value   # Column whose values will fill the new columns
  )

```

## Exploratory Data Analysis

Overview of Fertilizer Types

```{r}
strawb_fertilizers %>%
  count(Fertilizer, sort = TRUE)
```

Nitrogen and Potash appear as the most frequently reported fertilizers, suggesting they are the core components of conventional strawberry farming.

The graphs below show comparisons between conventional and organic strawberry production. The graph shows production over time in years (x-axis). It shows the number of acres harvested by state. The second graph shows the amount of strawberries produced per one hundred pounds.

```{r}
#| label: production graph conventional
#| warning: false
#| echo: false

production_data_c <- strawb_survey %>%
  filter(`Data Item` == "PRODUCTION, MEASURED IN CWT")

p1 <- ggplot(data = production_data_c, aes(x = Year, y = Value, color = State)) +
  geom_line(size = 1) +
  labs(
    title = "Conventional",
    x = "Year",
    y = "Production (per 100 pounds)",
    color = "State"
  ) +
  #facet_wrap(~ State, scales = "free_y", ncol = 2) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8) 
  ) 

```

```{r}
#| label: production graph organic
#| warning: false
#| echo: false

production_data_o <- strawb_organic %>%
  filter(`Data.Item` == "PRODUCTION, MEASURED IN CWT")

p2 <- ggplot(data = production_data_o, aes(x = Year, y = Value, color = State)) +
  geom_line(size = 1) +
  labs(
    title = "Organic",
    x = "Year",
    y = "",
    color = "State"
  ) +
  #facet_wrap(~ State, scales = "free_y", ncol = 2) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)
  ) 

```

```{r fig.align='center'}
#| label: production graphs combo
#| warning: false
#| echo: false

library(patchwork)

# Combine and add an overall title
(p1 | p2) + 
  plot_annotation(title = 'Strawberry Production by State',
                  theme = theme(plot.title = element_text(hjust = 0.35)))

```

Both of these graphs clearly shows us that California (in red) is the largest producer of conventional and organic strawberries. Florida (in yellow) is the second largest producer, but significantly behind California. The rest of the states have very low production compared to California and Florida.

**Geographic Patterns**

```{r}
library(ggplot2)
fert_by_state <- strawb_fertilizers_wide %>%
  group_by(State) %>%
  summarise(
    n_obs = sum(!is.na(`NITROGEN_ APPLICATIONS, MEASURED IN LB`)),  # count valid rows
    Nitrogen = sum(`NITROGEN_ APPLICATIONS, MEASURED IN LB`, na.rm = TRUE) / n_obs,
    Phosphate = sum(`PHOSPHATE_ APPLICATIONS, MEASURED IN LB`, na.rm = TRUE) / n_obs,
    Potash = sum(`POTASH_ APPLICATIONS, MEASURED IN LB`, na.rm = TRUE) / n_obs,
    Sulfur = sum(`SULFUR_ APPLICATIONS, MEASURED IN LB`, na.rm = TRUE) / n_obs
  ) %>%
  select(-n_obs) %>% 
  pivot_longer(-State, names_to = "Fertilizer", values_to = "Mean_Lb")


ggplot(fert_by_state, aes(x = State, y = Mean_Lb, fill = Fertilizer)) +
  geom_col(position = "dodge") +
  labs(title = "Average Fertilizer Use by State",
       y = "Average Pounds Applied", x = "State") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 6, angle = 45, hjust = 1)
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

The bar chart shows the average use of four fertilizer types—Nitrogen, Phosphate, Potash, and Sulfur—across major U.S. strawberry-producing states. California stands out as the dominant user by a very large margin, applying millions of pounds of each fertilizer, which reflects its role as the country’s largest strawberry producer by far. Florida ranks second, though at a much smaller scale, followed by Oregon, North Carolina, Washington, and New York. Nitrogen and Potash appear to be the most heavily used fertilizers across states, suggesting their key role in boosting plant growth and fruit production. Sulfur is used much less often, likely due to more specific soil needs. This pattern indicates that strawberry cultivation intensity and climate conditions vary significantly across regions, with California’s year-round production driving greater fertilizer demand.

Fertilizer plays a vital role in strawberry farming by supplying essential nutrients for plant growth, improving soil fertility, and increasing fruit yield and quality. Nitrogen promotes leafy growth and overall plant vigor but can cause environmental runoff issues if over-applied. Phosphate supports root development and early growth, though excessive use can lead to water pollution. Potash improves fruit size, sweetness, and disease resistance, while Sulfur helps regulate soil pH and supports nutrient uptake. Importantly, organic strawberry farms may also use fertilizers, but they must come from natural sources (e.g., compost, fish emulsion, mined minerals), whereas conventional farms can use synthetic fertilizers. This explains why some fertilizer categories appear across both systems, although organic volumes tend to be lower.

```{r, warning=FALSE}
strawb_fertilizers %>%
  group_by(Year, Fertilizer) %>%
  summarise(mean_value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = Year, y = mean_value, color = Fertilizer)) +
  geom_line(size = 1.2) +
  geom_point() +
  theme_minimal() +
  labs(title = "Fertilizer Use Over Time", y = "Average Use", x = "Year")
```

The line chart illustrates average fertilizer use in U.S. strawberry production from the late 1980s through 2023. Overall, fertilizer use generally increased from the 1990s to the mid-2010s, with Nitrogen and Potash consistently being the most heavily applied nutrients. These fertilizers are essential for strong plant growth, fruit size, and yield, which reflects the expanding commercial scale of strawberry farming over this period. Phosphate followed a similar but slightly lower trend, while Sulfur remained minimal until more recently, likely due to its use primarily for soil pH management and pest suppression rather than direct plant nutrition.

A striking feature of the graph is the sharp spike across all fertilizer types around 2018–2019, followed by a sudden drop by 2021. This pattern appears to be driven by the structure of the data set rather than true agronomic behavior specifically, those years include unusually high reported values from California, the nation’s top strawberry-producing state, paired with fewer reporting observations. The smaller denominator inflated the calculated average, creating an artificial spike. Once more states reported again or missing values returned, the average dropped back to typical levels. In reality, fertilizer practices tend to change gradually, so sudden jumps like this should be interpreted cautiously.

Beyond data artifacts, fertilizer use trends reflect evolving farming practices, higher yields demanded by consumers, and increased focus on soil health and environmental sustainability in recent years. The stabilization and slight decline after 2020 may relate to improved nutrient-efficiency programs, the rise of integrated pest and soil management, and increased interest in organic and reduced-input production.

The data set captures fertilizer usage primarily in conventional strawberry production system where the fertilizers are applied. In contrast, organic strawberry systems rely on nutrient source like composted manure or mineral sulfur rather than synthetic chemicals. This distinction matters because organic fertilizer release nutrients slowly and improve soil health, while conventional systems enable rapid growth but may pose environmental risks such as nitrate leaching and reduced soil biodiversity.

```{r, warning = False}
fert <- strawb_fertilizers %>%
  filter(Data.Item == " APPLICATIONS, MEASURED IN LB") %>%
  group_by(Year, State, Fertilizer) %>%
  summarise(total_value = sum(Value, na.rm = TRUE), .groups = "drop_last") %>%
  ungroup()
unique(strawb_fertilizers$Data.Item)


fert %>%
  ggplot(aes(x = Year, y = total_value, color = Fertilizer)) +
  geom_line(size = 1.2) +
  geom_point() +
  facet_wrap(~State, scales = "free_y") +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Set2") +
  theme_minimal() +
  labs(
    title = "Fertilizer Applications (lb) Over Time by State",
    y = "Total Fertilizer Use (lb)",
    x = "Year",
    color = "Fertilizer Type"
  ) +
  theme(plot.title = element_text(hjust = 0.5))
```

The graph above shows us the total fertilizer use by state, measured in pounds. The different colored lines represent the four fertilizer types. The graphs for North Carolina and New York have the shorted lines plotted. Furthermore we can see that California takes the lead in fertilizer use based on the values that lie on the y axis. There is a sharp increase in the usage of Potash around 2017/2018, with over 9 million pounds of it applied. 

**Conclusion**

This exploratory analysis provides an initial understanding of fertilizer use patterns in the US strawberry production:

-   Nitrogen and Potash dominate national usage

-   Fertilizer intensity is highest in major production states

-   Some fertilizers show correlated application patterns.

This report demonstrates the potential of fertilizer-use data to inform sustainability discussion in agriculture. However, a complete understanding of pros and cons requires linking fertilizer to other variables such as yield, soil quality, and economic return. Future analyses should incorporate these dimensions, along with spatial or weather data to assess fertilizer efficiency and environmental performance more directly. Future steps are our group continue with our analysis is to include linking fertilizer inputs to yield and environmental indicators to evaluate efficiency and sustainability discussions in agriculture.

Sources to be cited:

-   Yara: https://www.yara.com/knowledge-grows/five-important-facts-about-fertilizers/

-   Mayo Clinic: https://www.mayoclinic.org/healthy-lifestyle/nutrition-and-healthy-eating/in-depth/organic-food/art-20043880

-   UMN: https://extension.umn.edu/manage-soil-nutrients/quick-guide-fertilizing-plants#:\~:text=Plants%20lacking%20nitrogen%20will%20show,moisture%2C%20or%20damage %20to%20roots.
